---
title: "Decay functions"
output: rmarkdown::html_vignette
bibliography: '`r system.file("REFERENCES.bib", package = "accessibility")`'
vignette: >
  %\VignetteIndexEntry{Decay functions} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Some functions of the **accessibility** package, such as
`floating_catchment_area()` and `gravity()`, use decay functions to
continuously discount the weight of opportunities as travel costs become
larger. For convenience, the package ships with some of the most frequently
used functions in the accessibility literature, which are discussed below.
Additionally, users can pass custom functions to convert travel costs into
weights to be applied to the opportunities, which is discussed further down in
this the vignette.

# Currently available decay functions

## Binary

Also known as the step decay function, it's most commonly used in cumulative
opportunities measures.

$$
\begin{aligned}
  f(t_{ij})= \left\{
      \begin{array}{ll}
          1 & \quad \text{for }t_{ij} \leq T \\
          0 & \quad \text{for }t_{ij} > T
      \end{array}
    \right.\\
\end{aligned}
$$

Where $t_{ij}$ is the travel cost between origin *i* and destination *j*, and
$T$ is the travel cost cutoff.

## Linear

Weights decay linearly until the travel cost cutoff is reached. From this point
onward weights assume the value of 0.

$$
\begin{aligned}
  f(t_{ij})= \left\{
      \begin{array}{ll}
          (1 - t_{ij}/ T) & \quad \text{for }t_{ij} \leq T \\
          0 & \quad \text{for }t_{ij} > T
      \end{array}
    \right.\\
\end{aligned}
$$

Where $t_{ij}$ is the travel cost between origin *i* and destination *j*, and
$T$ is the travel cost cutoff.

## Negative exponential

$$
\begin{aligned}
  f(t_{ij})= e^{(-\beta t_{ij})}
\end{aligned}
$$

Where $t_{ij}$ is the travel cost between origin *i* and destination *j*, and
$\beta$ is the parameter that tells the speed of decay.

## Inverse power

$$
\begin{aligned}
  f(t_{ij})= \left\{
      \begin{array}{ll}
          1 & \quad \text{for } t_{ij}\leq 1 \\
          t_{ij}^{-\beta} & \quad \text{for }t_{ij} > 1
      \end{array}
    \right.\\
\end{aligned}
$$

Where $t_{ij}$ is the travel cost between origin *i* and destination *j*, and
$\beta$ is the parameter that tells the speed of decay.

## Stepped

Similar to the binary function, but can take an arbitrary number of steps. The
current implementation assumes that values changes at each step, instead of
right after it.

$$
\begin{aligned}
  f(t_{ij})= \left\{
      \begin{array}{ll}
          1 & \quad \text{for } t_{ij} \lt S_{1} \\
          v_{1} & \quad \text{for } t_{ij} \lt S_{2} \\
          v_{2} & \quad \text{for } t_{ij} \lt S_{3} \\
          ... \\
          v_{n-1} & \quad \text{for } t_{ij} \lt S_{n} \\
          v_{n} & \quad \text{otherwise} \\
      \end{array}
    \right.\\
\end{aligned}
$$

Where $t_{ij}$ is the travel cost between origin *i* and destination *j*, $n$
is the total number of steps, $S_{k}$ is the travel cost cutoff that delimits
the $k^{th}$ step, and $v_{k}$ is the value that the decay function assumes at
the $k^{th}$ step.

### Quick demonstration

All decay functions (`decay_*()`) take decay parameters as input and return a
function as output. This output function, in turn, takes a vector of numeric
values as input, the travel costs, and returns a list of numeric vectors as
output. Since the decay function can take multiple decay parameters as input,
each element of the output list refers to the opportunities weights calculated
with one of these decay parameters.

Let's check this behavior with an example. With the code snippet below, we
calculate the opportunities weights for the same travel costs, but using
different negative exponential decay values (0.2 and 0.3):

```{r}
library(accessibility)

output_fn <- decay_exponential(c(0.2, 0.3))

output_fn(c(10, 15, 20))
```

`decay_stepped()`, as the only decay function that takes more than one argument
(both `steps` and `weights`) names the output elements after the combination of
steps and weights:

```{r}
stepped_output <- decay_stepped(
  steps = list(c(10, 20, 30), c(10, 20, 30, 40)),
  weights = list(c(0.67, 0.33, 0), c(0.75, 0.5, 0.25, 0))
)

stepped_output(c(15, 25, 35, 45))
```

With the code below, we demonstrate each decay function with travel costs
ranging from 1 to 100:

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), out.width = "80%", fig.width = 6, fig.height = 6}
library(data.table)
library(ggplot2)

binary <- decay_binary(cutoff = 50)
linear <- decay_linear(cutoff = 50)
negative_exp <- decay_exponential(decay_value = 0.2)
inverse_power <- decay_power(decay_value = 0.2)
stepped <- decay_stepped(steps = c(30, 60, 90), weights = c(0.67, 0.33, 0))

travel_costs <- seq(1, 100, 0.1)

weights <- data.table(
  minutes = travel_costs,
  binary = as.numeric(binary(travel_costs)[["50"]]),
  linear = linear(travel_costs)[["50"]],
  negative_exp = negative_exp(travel_costs)[["0.2"]],
  inverse_power = inverse_power(travel_costs)[["0.2"]],
  stepped = stepped(travel_costs)[["s(30,60,90);w(0.67,0.33,0)"]]
)

# reshape data to long format
weights <- melt(
  weights,
  id.vars = "minutes",
  variable.name = "decay_function",
  value.name = "weights"
)

ggplot(weights) +
  geom_line(
    aes(minutes, weights, color = decay_function),
    show.legend = FALSE
  ) +
  facet_wrap(. ~ decay_function, ncol = 2) +
  theme_minimal()
```

# Using a custom decay function

**accessibility** also allows you to use a custom decay function, instead of one
of the functions shipped with the package. A valid decay function is one that
takes a `numeric` vector of travel costs as input and returns either:

- A `numeric` vector of weights, with the same length of input, or;
- A named `list` of `numeric` vectors to be used as weights, each one with the
  same length of input.

Let's check the difference between each case with an example. Suppose we want to
use a very simple decay function that defines the weights as the multiplicative
inverse of travel cost - i.e. $travel\_cost^{-1}$. We just have to take care of
the case when travel cost is less than 1, in which case the function would
return values greater than 1 and which we will replace with 1, but otherwise the
implementation is pretty simple:

```{r}
my_decay <- function(travel_cost) {
  weights <- 1 / travel_cost
  weights[weights > 1] <- 1
  return(weights)
}
```

Given a `numeric` vector of travel costs, the function returns a `numeric`
vector of weights:

```{r}
my_decay(c(0, 0.5, 1, 2, 5, 10))
```

Using this function to calculate accessibility is as easy as any of the built-in
decay functions:

```{r}
data_dir <- system.file("extdata", package = "accessibility")

travel_matrix <- readRDS(file.path(data_dir, "travel_matrix.rds"))
land_use_data <- readRDS(file.path(data_dir, "land_use_data.rds"))

custom_gravity <- gravity(
  travel_matrix,
  land_use_data,
  opportunity = "jobs",
  travel_cost = "travel_time",
  decay_function = my_decay
)
head(custom_gravity)
```

Great! But now suppose we want to change this function a bit. Instead of the
simple multiplicative inverse of the travel cost, we want to multiply this
inverse by a given decay parameter. Now we face a "problem": our function would
need to take two inputs (the travel cost, as above, and the decay parameter to
multiply the travel cost), but `gravity()` can only take functions that receive
a single input (the travel cost). In this case, we resort to the same strategy
used in the decay functions shipped with the package, we create a function that
takes a decay parameter as input and returns a function that takes travel cost
input as output:

```{r}
my_second_decay <- function(decay_parameter) {
  function(travel_cost) {
    weights <- 1 / (decay_parameter * travel_cost)
    weights[weights > 1] <- 1
    return(weights)
  }
}

output_fn <- my_second_decay(2)
output_fn(c(0, 0.5, 1, 2, 5, 10))

# compare to the first custom decay function
my_decay(c(0, 0.5, 1, 2, 5, 10))
```

Great, it works! In fact, we can achieve the exact same `gravity()` result shown
above if we use `my_second_decay(1)`, instead of `my_decay`:

```{r}
second_custom_gravity <- gravity(
  travel_matrix,
  land_use_data,
  opportunity = "jobs",
  travel_cost = "travel_time",
  decay_function = my_second_decay(1)
)
head(second_custom_gravity)
```

A small difference is that in the first example we passed `my_decay` (the
function *object*) to the `decay_function` parameter, whereas in the second we
passed the function *call* `my_second_decay(1)`. That's because
`my_second_decay()` (and the built-in decay functions as well) is actually a
function "factory": it's a function that returns a function.

```{r}
decay_power(1)
```

What if we want our custom function to take many decay parameters as inputs,
just like the built-in decay functions? In this case, our output function has to
return a `list` of `numeric` weights named after the decay parameters we have
set:

```{r}
my_third_decay <- function(decay_parameter) {
  function(travel_cost) {
    weighting_list <- lapply(
      decay_parameter,
      function(x) {
        weights <- 1 / (x * travel_cost)
        weights[weights > 1] <- 1
        return(weights)
      }
    )

    names(weighting_list) <- decay_parameter
    weighting_list
  }
}

output_fn <- my_third_decay(c(1, 2))
output_fn(c(0, 0.5, 1, 2, 5, 10))

# compare to the first and second custom decay functions

my_decay(c(0, 0.5, 1, 2, 5, 10))

my_second_decay(2)(c(0, 0.5, 1, 2, 5, 10))
```

This new custom decay function is a bit more complex than the previous two, yet
much more powerful. We can use it in conjunction with `gravity()` (and the other
accessibility functions that take decay functions as input as well) to calculate
the accessibility with multiple decay parameters in a single call:

```{r}
third_custom_gravity <- gravity(
  travel_matrix,
  land_use_data,
  opportunity = "jobs",
  travel_cost = "travel_time",
  decay_function = my_third_decay(c(1, 2))
)
third_custom_gravity
```

When we create our own custom decay function, we don't have to think about
general cases, just the cases that matter for us. Creating a simple function
that converts one vector into another, therefore, might be enough, given our
needs. If, however, you're testing a new decay function and you want to check
its sensitivity to different decay parameters, for example, you might be better
off developing a decay function that can take many decay values as inputs and
returns a list of weight vectors as output.
