---
title: 'Decay functions'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: "references.bib"
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Decay functions} 
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8]{inputenc}
bibliography: references.json
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# 1. Introduction

Some functions of the `accessibility` package, such as `floating_catchment_area()` and `gravity_access()`, use decay functions to continuously discount the weight of opportunities as travel costs become larger. For convenience, the package currently includes four decay functions discussed in [@kwan1998space]. See the implemented functions below. Additionally, the package also allows users to pass any custom decay function that converts travel cost `t_ij` to an impedance factor (see section 3).

# 2. Currently available decay functions


## 2.1 Binary decay function

Also known as 'step' decay function, commonly used in cumulative opportunity measures.

$$
\begin{aligned}
  f(t_{ij})= \left\{
      \begin{array}{ll}
          1 & \quad \text{for }t_{ij} \leq T \\
          0 & \quad \text{for }t_{ij} > T
      \end{array}
    \right.\\
\end{aligned}
$$

Where $t_{ij}$ is the travel cost between origin *i* and destination *j*; and $T$ is the `cutoff` of maximum travel cost.

## 2.2 Linear decay function

$$
\begin{aligned}
  f(t_{ij})= \left\{
      \begin{array}{ll}
          (1 - t_{ij}/ T) & \quad \text{for }t_{ij} \leq T \\
          0 & \quad \text{for }t_{ij} > T
      \end{array}
    \right.\\
\end{aligned}
$$

Where $t_{ij}$ is the travel cost between origin *i* and destination *j*; and $T$ is the `cutoff` of maximum travel cost.


## 2.3 Negative exponential.



$$
\begin{aligned}
  f(t_{ij})= e^{(-\beta t_{ij})}
\end{aligned}
$$

Where: $t_{ij}$ is the travel cost between origin *i* and destination *j*; and $\beta$ is the `decay_value` that tells the speed of decay.


## 2.4 Inverse power

$$
\begin{aligned}
  f(t_{ij})= \left\{
      \begin{array}{ll}
          1 & \quad \text{for } t_{ij}\leq 1 \\
          t_{ij}^{-\beta} & \quad \text{for }t_{ij} > 1
      \end{array}
    \right.\\
\end{aligned}
$$

Where $t_{ij}$ is the travel cost between origin *i* and destination *j*; and $\beta$ is the `decay_value` that tells the speed of decay.



#### Quick demonstration of how these decay functions look like

```{r}
library(accessibility)
library(data.table)
library(ggplot2)

# Generate inputs
vec <- 0:100
decay_value <- 0.2
cutoff <- 50

# Return functions
step        <- decay_binary(cutoff=cutoff)
linear      <- decay_linear(cutoff=cutoff)
exponential <- decay_exponential(decay_value=decay_value)
power       <- decay_power(decay_value = decay_value)

df <- data.table(minutes       = vec,
                 binary        = step(vec),
                 linear        = linear(vec),
                 exponential   = exponential(vec),
                 inverse_power = power(vec)
                 )

# reshape the data to long format
df2 <- data.table::melt.data.table(data = df, 
                                   id.vars = 'minutes', 
                                   variable.name = 'decay_function', 
                                   value.name = 'impedance_factor')

# plot
ggplot() +
  geom_line(data=df2, aes(x=minutes, y=impedance_factor, color=decay_function), show.legend = FALSE) +
  facet_wrap(.~decay_function, ncol = 2) +
  theme_minimal()

```



# 3. Using a custom decay function

Let's use the sample data available in the `accessibility` package o demonstrate how you could use a custom decay function. Let's load the data first.

```{r, message=FALSE, warning=FALSE}

# load input data
data_path <- system.file("extdata/ttm_bho.rds", package = "accessibility")
ttm <- readRDS(data_path)
head(ttm)

```

Now let's create a very simple custom decay function that uses a parameter `k` to convert the travel cost `t_ij` between origin `i` and destination `j` into an impedance factor.


```{r}
# create custom decay function
custom_decay_fun <- function(k) {
  impedance <- function(t_ij) {
    f <- 1 / t_ij ^ k
    return(f)
  }
}
```
You can use your `custom_decay_fun()` inside the `floating_catchment_area()` and `gravity_access()` functions like this:

```{r, message = FALSE}
# calculate gravity-based accessibility
grav_custom <- gravity_access(
                  data = ttm,
                  opportunity_col = 'jobs',
                  decay_function = custom_decay_fun(k=0.5),
                  travel_cost_col = 'travel_time',
                  by_col = 'from_id'
                  )

head(grav_custom)
```
## References
