---
title: "Decay functions"
output: rmarkdown::html_vignette
bibliography: '`r system.file("REFERENCES.bib", package = "accessibility")`'
vignette: >
  %\VignetteIndexEntry{Decay functions} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Some functions of the **accessibility** package, such as
`floating_catchment_area()` and `gravity()`, use decay functions to
continuously discount the weight of opportunities as travel costs become
larger. For convenience, the package ships with some of the most frequently
used functions in the accessibility literature, which are discussed below.
Additionally, users can pass custom functions to convert travel costs into
weights to be applied the opportunities, which is discussed further down the
vignette.

# Currently available decay functions

## Binary

Also known as the step decay function, it's most commonly used in cumulative
opportunities measures.

$$
\begin{aligned}
  f(t_{ij})= \left\{
      \begin{array}{ll}
          1 & \quad \text{for }t_{ij} \leq T \\
          0 & \quad \text{for }t_{ij} > T
      \end{array}
    \right.\\
\end{aligned}
$$

Where $t_{ij}$ is the travel cost between origin *i* and destination *j*, and
$T$ is the travel cost cutoff.

## Linear

Weights decay linearly until the travel cost cutoff is reached. From this point
onward weights assume the value of 0.

$$
\begin{aligned}
  f(t_{ij})= \left\{
      \begin{array}{ll}
          (1 - t_{ij}/ T) & \quad \text{for }t_{ij} \leq T \\
          0 & \quad \text{for }t_{ij} > T
      \end{array}
    \right.\\
\end{aligned}
$$

Where $t_{ij}$ is the travel cost between origin *i* and destination *j*, and
$T$ is the travel cost cutoff.

## Negative exponential

$$
\begin{aligned}
  f(t_{ij})= e^{(-\beta t_{ij})}
\end{aligned}
$$

Where $t_{ij}$ is the travel cost between origin *i* and destination *j*, and
$\beta$ is the parameter that tells the speed of decay.

## Inverse power

$$
\begin{aligned}
  f(t_{ij})= \left\{
      \begin{array}{ll}
          1 & \quad \text{for } t_{ij}\leq 1 \\
          t_{ij}^{-\beta} & \quad \text{for }t_{ij} > 1
      \end{array}
    \right.\\
\end{aligned}
$$

Where $t_{ij}$ is the travel cost between origin *i* and destination *j*, and
$\beta$ is the parameter that tells the speed of decay.

## Stepped

Similar to the binary function, but can take an arbitrary number of steps. The
current implementation assumes that values changes at each step, instead of
right after it.

$$
\begin{aligned}
  f(t_{ij})= \left\{
      \begin{array}{ll}
          1 & \quad \text{for } t_{ij} \lt S_{1} \\
          v_{1} & \quad \text{for } t_{ij} \lt S_{2} \\
          v_{2} & \quad \text{for } t_{ij} \lt S_{3} \\
          ... \\
          v_{n-1} & \quad \text{for } t_{ij} \lt S_{n} \\
          v_{n} & \quad \text{otherwise} \\
      \end{array}
    \right.\\
\end{aligned}
$$

Where $t_{ij}$ is the travel cost between origin *i* and destination *j*, $n$
is the total number of steps, $S_{k}$ is the travel cost cutoff that delimits
the $k^{th}$ step, and $v_{k}$ is the value that the decay function assumes at
the $k^{th}$ step.

### Quick demonstration

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), out.width = "80%", fig.width = 6, fig.height = 6}
library(accessibility)
library(data.table)
library(ggplot2)

binary <- decay_binary(cutoff = 50)
linear <- decay_linear(cutoff = 50)
negative_exp <- decay_exponential(decay_value = 0.2)
inverse_power <- decay_power(decay_value = 0.2)
stepped <- decay_stepped(steps = c(30, 60, 90), weights = c(0.66, 0.33, 0))

travel_costs <- seq(1, 100, 0.1)

weights <- data.table(
  minutes = travel_costs,
  binary = as.numeric(binary(travel_costs)),
  linear = linear(travel_costs),
  negative_exp = negative_exp(travel_costs),
  inverse_power = inverse_power(travel_costs),
  stepped = stepped(travel_costs)
)

# reshape data to long format
weights <- melt(
  weights,
  id.vars = "minutes",
  variable.name = "decay_function",
  value.name = "weights"
)

ggplot(weights) +
  geom_line(
    aes(minutes, weights, color = decay_function),
    show.legend = FALSE
  ) +
  facet_wrap(. ~ decay_function, ncol = 2) +
  theme_minimal()
```

# Using a custom decay function

Any function that takes a `numeric` vector of travel costs and returns a
`numeric` vector, with the same length of input, of weights to be applied to
the opportunities can be considered a decay function.

Let's say that we wanted to use a very simple decay function that defines the
weights as the multiplicative inverse of travel cost - i.e.
$travel\_cost^{-1}$. We just have to take care of the case when travel cost is
less than 1, in which case the function would return values greater than 1 and
which we will replace with 1, but otherwise the implementation is pretty
simple:

```{r}
my_decay <- function(travel_cost) {
  weights <- 1 / travel_cost
  weights[weights > 1] <- 1
  return(weights)
}
```

Given a `numeric` vector of travel costs, the function returns a `numeric`
vector of weights:

```{r}
my_decay(c(0, 0.5, 1, 2, 5, 10))
```

Using this function to calculate accessibility is as easy as any of the built-in
decay functions:

```{r}
data_dir <- system.file("extdata", package = "accessibility")

travel_matrix <- readRDS(file.path(data_dir, "travel_matrix.rds"))
land_use_data <- readRDS(file.path(data_dir, "land_use_data.rds"))

custom_gravity <- gravity(
  travel_matrix,
  land_use_data,
  opportunity = "jobs",
  travel_cost = "travel_time",
  decay_function = my_decay
)
head(custom_gravity)
```

In fact, we could achieve the exact same result as we did with our custom
function using `decay_power(1)`:

```{r}
power_gravity <- gravity(
  travel_matrix,
  land_use_data,
  opportunity = "jobs",
  travel_cost = "travel_time",
  decay_function = decay_power(1)
)
head(power_gravity)
```

A small difference is that in the first example we passed `my_decay` (the
function *object*) to the `decay_function` parameter, whereas in the second we
passed the function *call* `decay_power(1)`. That's because `decay_power()`
(and the other built-in decay functions as well) is actually a function
"factory": it's a function that returns a function, as we can see below:

```{r}
decay_power(1)
```

This is because the built-in functions aim to be as generally used as possible,
so we have to able to use them with different parameters, cutoffs, etc. In
practice, every time we call these function we create another function that
calculates travel costs exactly as we want.

When we create our own custom decay function, we don't have to think about
general cases, just the cases that matter for us. Creating a simple function
that converts one vector into another, therefore, suffices our needs. In both
examples above, therefore, we're passing a function to `decay_function` - one
we created ourselves and another created by `decay_power()`.
