---
title: "Calculating accessibility inequality and poverty"
output: rmarkdown::html_vignette
bibliography: '`r system.file("REFERENCES.bib", package = "accessibility")`'
vignette: >
  %\VignetteIndexEntry{Calculating accessibility inequality and poverty} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**accessibility** not only includes functions to calculate accessibility levels
from travel time matrices and land use datasets, but also to calculate
accessibility inequality and poverty from an accessibility distribution. This
vignette briefly presents these functions.

# Demonstration on sample data

We first need to calculate the accessibility levels that we are going to use in
this demonstration. To do so, we use the `cumulative_cutoff()` function and the
sample data included in the package.

```{r}
library(accessibility)

data_dir <- system.file("extdata", package = "accessibility")
travel_matrix <- readRDS(file.path(data_dir, "travel_matrix.rds"))
land_use_data <- readRDS(file.path(data_dir, "land_use_data.rds"))

access <- cumulative_cutoff(
  travel_matrix,
  land_use_data,
  opportunity = "jobs",
  travel_cost = "travel_time",
  cutoff = 30
)
head(access)
```

The functions we’ll be demonstrating in this section take much of the same
information as input, including:

- the accessibility distribution;
- a sociodemographic dataset containing information such as the distribution of
  people in the study area and their income;
- the name of the column in the accessibility dataset with the opportunities
  whose accessibility should be considered in the calculation; and
- the name of the column in the sociodemographic dataset with the number of
  people in each cell, used to weigh accessibility levels when calculating
  inequality/poverty.

## Palma Ratio

`palma_ratio()` calculates the Palma Ratio of a given accessibility
distribution. Originally defined as the income share of the wealthiest 10% of a
population divided by the income share of the poorest 40%, this measure has been
adapted in transport planning as the average accessibility of the wealthiest 10%
divided by the average accessibility of the poorest 40%.

This function includes an additional `income` parameter, used to list the column
in the sociodemographic dataset with the income variable that should be used to
classify the population in socioeconomic groups. Please note that this variable
should describe income per capita (e.g. mean income per capita, household income
per capita, etc), instead of the total amount of income in each cell.

```{r}
palma <- palma_ratio(
  access,
  sociodemographic_data = land_use_data,
  opportunity = "jobs",
  population = "population",
  income = "income_per_capita"
)
palma
```

## Gini Index

`gini_index()` calculates the Gini Index of a given accessibility distribution.

```{r}
gini <- gini_index(
  access,
  sociodemographic_data = land_use_data,
  opportunity = "jobs",
  population = "population"
)
gini
```

## Foster-Greer-Thorbecke (FGT) poverty measures

`fgt_poverty()` calculates the FGT metrics, a family of poverty measures
originally proposed by @foster1984class that capture the extent and severity of
poverty within an accessibility distribution. The FGT family is composed of
three measures that differ based on the $\alpha$ parameter used to calculate
them (either 0, 1 or 2) and which also changes their interpretation:

- with $\alpha = 0$ (FGT0) the measure captures the extent of poverty as a
  simple headcount - i.e. the proportion of people below the poverty line;
- with $\alpha = 1$ (FGT1) the measure, also know as the "poverty gap index",
  captures the severity of poverty as the average percentage distance between
  the poverty line and the accessibility of individuals below the poverty line;
- with $\alpha = 2$ (FGT2) the measure simultaneously captures the extent and
  the severity of poverty by calculating the number of people below the poverty
  line weighted by the size of the accessibility shortfall relative to the
  poverty line.

This function includes an additional `poverty_line` parameter, used to define
the poverty line below which individuals are considered to be in accessibility
poverty.

```{r}
poverty <- fgt_poverty(
  access,
  opportunity = "jobs",
  sociodemographic_data = land_use_data,
  population = "population",
  poverty_line = 95368
)
poverty
```

# References
